Trail Saving on Backtrack

### 2020-05-11 figures/scatter2.tex
final version

\documentclass{standalone}
\usepackage{tikz,pgfplotstable,pgfplots}
\pgfplotsset{compat=1.13}
\begin{document}
\begin{tikzpicture}
\begin{axis}[title=Propagations Per Second in Cadical,axis equal,ymin=0,xmin=0,
width=11cm,height=11cm,xlabel=Baseline props/sec (in M),ylabel=Trail Saving props/sec (in M),axis x line=bottom,axis y line=left,xlabel near ticks,ylabel near ticks]
\addplot[only marks,mark size=1pt,mark=*,mark options={fill=black}]
	table[x=original,y=trail]
{scatter.txt};
\addplot [red,samples at={0,1}] {x};
\draw [blue,dashed] (rel axis cs:0,0) -- (rel axis cs:1,1);
\draw (axis cs:2,6) node[text width=3cm,align=center] {213 instances had increase in props/sec by $>$ 0.1M};
\draw (axis cs:6,2) node[text width=3cm,align=center] {149 instances had decrease in props/sec by $>$ 0.1M};
\end{axis}
\end{tikzpicture}
\end{document}

>>>>

\documentclass{standalone}
\usepackage{tikz,pgfplotstable,pgfplots}
\pgfplotsset{compat=1.13}
\begin{document}
\begin{tikzpicture}
\begin{axis}[title=Baseline vs Trail Saving Running Times in Cadical,axis equal,ymin=0,xmin=0,
width=11cm,height=11cm,xlabel=Baseline running time (sec),ylabel=Trail Saving running time (sec),axis x line=bottom,axis y line=left,xlabel near ticks,ylabel near ticks]
\addplot[only marks,mark size=1pt,mark=*,mark options={fill=black}]
	table[x=original,y=trail]
{scatter2.txt};
\addplot [red,samples at={0,1}] {x};
\draw [blue,dashed] (rel axis cs:0,0) -- (rel axis cs:1,1);

\end{axis}
\end{tikzpicture}
\end{document}

### 2020-05-11 paper.tex
final version

\title{Trail Saving on Backtrack}
\author{Randy Hickey \and Fahiem Bacchus}
\institute{Department of Computer Science, University of Toronto, Canada\\
  \email{rgh000@gmail.com, fbacchus@cs.toronto.edu}}

@@

In this paper we present an new trail saving method whereby we save
the backtracked part of the solver's trail and attempt to use that
information to make the solver's redescent more efficient. Unlike
\cbt, our trail saving method preserves the traditional invariants of
the SAT solver and its basic version is very simple to implement. It
allows the search to retain complete control over the order of
decisions, but helps make propagation faster. We develop some
enhancements to make the idea more effective, and demonstrate
experimentally that it performs as well as and often better than
chronological backtracking. We also show that with our enhancements we
are able to improve the performance of state-of-the-art solvers.

@@

This means that the decision levels on the trail are no longer
contiguous, as $\deepestLit$ has a different level than the other
literals at level $j$ (if $j\neq \btL$). This change has a number of
consequences for the \sat solver's operation, all of which were
described in \cite{DBLP:conf/lpar/JiangZ13}. M{\"{o}}hle and Biere
\cite{DBLP:conf/satt/MohleB19} showed that despite these consequences
partial backtracking can be made to preserve the soundness of a CDCL
solver.

@@

The second impact, changing variable scores, is partially addressed in
\cite{DBLP:conf/sat/MohleB19} who utilize the ideas of
\cite{DBLP:journals/jsat/TakRH11} to backtrack to a level where the
decisions would be unchanged. However, if the length of the backtrack
is greater than $100$ there could still be a divergence between the
variable decisions generated in standard backtracking and \cbt. An
argument is also given in \cite{DBLP:journals/jsat/TakRH11} that the
third impact, changing literal reasons, is not significant. However,
the experiments in \cite{DBLP:journals/jsat/TakRH11} were run before
good notions of clause quality were known
\cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results indicate
that once clause quality is accounted, for changing the literal
reasons can have a significant impact.

@@

% \note{
%   \begin{enumerate}
%   \item Use simple backtrack and save to motivate and state the
%       invariant. 
%   \item Introduce notation for concatenating two trails.
%   \item Name and state the invariant in text (e.g. a
%       defn)---concenating the cached trail with the current trail
%       yields a trail that is always reason valid, but not necessarily
%       non-contradictory, non-conflicting, nor propagation complete).
%   \item Proof (simple) that the invariant holds for caching on
%       backtrack.
%   \item On backtrack at least the asserted literal and any new
%       propagants at the backtrack level are added.
%   \item Prove that the invariant holds (???do we need an invariant or
%       simply a property of the augmented trail) for trails with extras
%       inbeween.
%       $T_{\mathit{old}} + T_{\mathit{newlyAdded}} +
%       T_{\mathit{saved}}$ also satisfies the invariant
%   \item re-acheving propagation completeness for the augmented trail
%       (a) only add one level of the cached trail at a time, and do
%       unit prop after (b) only add non-contradicted decision from the
%       old trail. Also note that conflicts are valid and can be learnt
%       from even if the augmented trail is not propagation complete. 
%   \item Finally figure out a way to formalize the extra technique of
%       storing on top after repeated backtracking. 
%   \item Explain how freedom for the search is maintained by makeing
%       decisions then consulting the trail.
%   \item 
%   \end{enumerate}
% }

@@

% \section{Related Work}
% \subsection{Trail Saving for Assumption-Based SAT}
% Trail saving was briefly introduced in the context of assumption-based
% SAT \cite{DBLP:conf/sat/HickeyB19} in order to speed up the
% propagation of the assumptions each time the solver backtracked past
% the assumption levels. Note that the process is the same, but it only
% works to restore the propagants and implicants of the assumptions and
% only after a learnt unit clause. Becuase the number of propagants and
% implicants of the assumptions tends to become dominated by the number
% of assumptions in the applications it was tested on, the technique did
% not yield a significant improvement. It would be interesting to
% combine the extended version of trail saving outlined in this paper
% with trail saving over the assumptions to see if that could produce
% better results.

% \subsection{Comparison to Chronological Backtracking}
% As mentioned earlier, chronological backtracking also achieves some of
% the theoretical speedups that trail saving achieves. One advantage of
% chronological backtracking is that it does not require any caching of
% the trail or accessing such a cache during propagation; it simply
% leaves a vast portion of the trail intact rather than backtracking all
% the way to the conflict level. Chronological backtracking will also
% speed up propagations slightly more than trail saving, because it
% leaves more literals assigned at once which leads to clauses being
% detected as unit more quickly.

% One disadvantage that chronological backtracking has compared to trail
% saving is that it doesn't allow the search as much flexibility to jump
% out of local neighbourhoods, as the common variable selection
% heuristics tend to depend on
% \cite{DBLP:conf/dac/MoskewiczMZZM01,DBLP:conf/sat/2015,DBLP:conf/sat/LiangGPC16}. Because
% chronological backtracking leaves a large portion of the trail
% assigned, it will behave similarly to how the solver would behave if
% it was to force the decisions that do occur within this large portion
% of literals, whereas trail saving does not force any decisions. As
% variable selection heuristics become more refined, it is likely that
% adhering to them will become more crucial to performance.

% Another disadvantage is that, because propagation is not performed at
% every decision level, literals that remain assigned during
% chronological backtracking (but otherwise would have been unassigned)
% do not have a chance to move their decision levels up. An example of
% this is when the solver was supposed to backtrack over a literal that
% would end up being unit propagated at level 0, but stays assigned at
% its current decision level when chronological backtracking is
% used. This could potentially effect the lbd score or length of the
% learnt clause that gets generated at the next conflict. Additionally,
% chronological backtracking breaks the invarant that all literals
% appear on the trail in monotonically increasing order by decision
% level, which makes its implementation slightly more complicated.

@@

In Figure~\ref{fig:cadicalNew} we used the newest version of cadical
(downloaded as of January 1, 2020) as the baseline solver, in
Figure~\ref{fig:cadicalOld} we used the version of cadical published
in \cite{DBLP:conf/sat/MohleB19} as the baseline, and in
Figure~\ref{fig:maple} we used MapleLCMDist \cite{DBLP:conf/ijcai/LuoLXML17,xiao2017maplelrb} as
the baseline. Each of the baselines were run with standard
non-chronological backtracking. We then refer to versions of each
solver with additional features implemented on top by adding
suffixes. "-chrono" refers to the solver with \cbt enabled (using the
solver's default settings), "-trail" refers to the baseline with plain
trail saving added (as described in Figure~\ref{fig:trailUse}),
"-trail-multipleBT" refers to the baseline with trail saving plus the
first enhancement of saving over multiple backtracks,
"-trail-multipleBT-lookahead" also adds the enhancement of lookahead
for conflicts by 2 decision levels, and
"-trail-multipleBT-lookahead-reason" also adds the final enhancement
to cease trail saving once a reason of "low quality" is reached. For
more details on the enhancements, please see section 4.2.

@@

\begin{figure}[H]
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      cadical                  & 532          & 314 &  218  & 4025                             \\ \hline
      cadical-chrono          & 525          & 308 &  217  & 4086                             \\ \hline
      cadical-trail           & 529          & 310 &  \textbf{219}  & 4060                   \\ \hline
      cadical-trail-multipleBT & 530 & 313 & 217 & 3930 \\ \hline
      cadical-trail-multipleBT-lookahead & 531          & 313 &  218  & 4020                      \\ \hline
      cadical-trail-multipleBT-lookahead-reason & \textbf{538} & \textbf{319} & \textbf{219} & \textbf{3854} \\ \hline
    \end{tabular}
    \caption{Table of results for cadical, version pulled from github as of January 1, 2020.}
    \label{fig:cadicalNew}
\end{figure}

@@

In Figure~\ref{fig:props} we plotted the propagations per second for
each instance with and without trail saving enabled in cadical (rounded
to the nearest 0.1 million). Only
the 506 instances solved by both solvers are plotted. With trail saving
enabled, 213 instances had more propagations per second, 149 instances
had less propagations per second, and 144 instances had about the same
number of propagations per second. On the graph, points above the diagonal
represent instances where trail saving had faster propagation. The
graph illustrates that the majority of the time
cadical with trail saving enabled propagated literals faster than
cadical without trail saving enabled, as we predicted would be the
case earlier in this paper.

\begin{figure}\centering\includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\label{fig:props}\end{figure}

@@

%\section{Future Work}

%With more clever data structures the efficiency of restoring the trail
%could be improved. There are a number of parameters which can be
%tuned, including how often to refresh the trail, how large the cutoff
%point for rejecting "long" reasons should be, how long to wait to
%prune the trail cache if one is appending multiple trail caches
%together, how far to look ahead on the trail cache for conflicts (in
%terms of number of decisions or total number of literals). We only
%experimented with a few different settings of these parameters that
%made sense, but more careful fine-tuning of these parameters could
%further enhance the effectiveness of the technique. There are
%potentially other techniques one could use a trail cache for,
%including inprocessing steps that involve probing. A look down the
%trail cache is an incomplete but very fast way to do a probing step,
%as opposed to propagation. It is also possible that trail saving and
%chronological backtracking could be combined in some way to produce
%even better results.

@@

\iffalse \textit{Example 1} Suppose the literal $y$ can be restored
from the trail cache, and its reason clause $cr$ contains the literals
${y \lnot x_1, \lnot x_2, ..., \lnot x_n}$. This means that
$x1, ..., x_n$ have become true and are currently on the
trail. Suppose $x_n$ is the most recent literal on the trail and
Without trail saving, the propagation routine will have to access $cr$
from the watchlist of one of the literals $\blacksquare$\newline

\textit{Example 1} Suppose the literals $x_1, ... x_n$ are on the
trail cache, with none of them having a null reason, and $x_1$ becomes
true (e.g., the solver makes a decision for $x_1$ to be true). Then,
$x_2, ... x_n$ are implied by the current trail. Now we will examine
the propagation of a clause c which consists of the literals
$l, \lnot x_1, \lnot x_2, ..., \lnot x_n$ with and without trail
saving. Assume $l$ and $\lnot x_1$ are the current watched literals.

Without trail saving, $\lnot x_1$ becomes false and a new watcher is
searched for and found 1 literal later, $\lnot x_2$, and then
$\lnot x_1$ and $\lnot x_2$ are swapped. Next $\lnot x_2$ becomes
false, a new watcher is searched for and found 2 literals later,
$\lnot x_3$, and then $\lnot x_2$ and $\lnot x_3$ are swapped. This
process will continue until all $x_n$ literals are false, c becomes
unit, and $l$ becomes forced to be true. This propagation took
$1 + 2 + ... + n$ searches through clause c, for a total of $O(n^2)$
searches.

With trail saving, $x2, ... x_n$ are all assigned to be true as soon
as $x_1$ is made true, so immediately the values of literals
$\lnot x_1, ..., \lnot x_n$ in c are false. During propagation, a new
watcher is searched for to replace $\lnot x_2$, n literals are
searched over without finding any, thus c is detected as unit and $l$
becomes forced to be true. This propagation took $O(n)$ searches
through clause c. $\blacksquare$\newline \fi

\iffalse
\begin{figure}\includegraphics[scale=0.8]{figures/cactus_cadical.pdf}\caption{\small{Comparison
        of run times for versions of cadical. cadical is without
        chronological backtracking, cadical-chrono is with
        chronological backtracking, cadical-trail is with the first
        version of trail saving, cadical-trail-2levels-ahead is the
        trail saving with all enhancements added.}}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{figures/cactus_maple2.pdf}\caption{\small{Comparison
        of run times for versions of MapleSAT. maplelcmdist is from
        the 2017 SAT Competition, maplelecmdist-chrono is with
        chronological backtracking (2018 SAT Competition),
        maplelcmdist-trail is with the first version of trail saving,
        maplelcmdist-trail-enhanced is with trail saving and all
        enhancements added.}}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{figures/test.pdf}\caption{}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{figures/test2.pdf}\caption{}\end{figure}
\fi
\iffalse
\begin{figure}\includegraphics[scale=0.8]{figures/cactus_maple.pdf}\caption{\small{Comparison
        of run times for versions of MapleSAT. maplelcmdist is from
        the 2017 SAT Competition, maplelecmdist-chrono is with
        chronological backtracking (2018 SAT Competition),
        maplelcmdist-trail is with the first version of trail saving,
        maplelcmdist-chrono-dl is with chronological backtracking and
        duplicate learnts (2019 SAT Competition),
        maplelcmdist-trail-dl is with trail saving and duplicate
        learnts}}\end{figure}
\fi

\iffalse
\begin{figure}[t]
    \begin{center}
        \fss{9pt}{10pt}
        \setlength\tabcolsep{1pt}
        \begin{tabular}{| p{4.2cm} | c | c | c | c | c | c | c | c |}
          \hline
          & \multicolumn{4}{|c|}{Total} & \multicolumn{2}{|c|}{Unweighted} & \multicolumn{2}{|c|}{Weighted} \\ \hline
          & MaxHS & +/- & \hspace{1pt} RC2 \hspace{1pt} & +/- & MaxHS & \hspace{1pt} RC2 \hspace{1pt} & MaxHS & \hspace{1pt} RC2 \hspace{1pt} \\ \hline
          original & 6052 & 0/0 & 6030 & 0/0 & 3940 & 3993 & 2112 & 2037 \\ \hline
          enqueue assumptions as set & 6131 & 114/35 & 6081 & \hspace{0.5pt} 99/48 \hspace{0.5pt} & 3995 & 4032 & 2136 & 2049 \\ \hline
          enqueue assumptions as set + \newline save literals after learnt units & 6136 & 120/36 & 6079 & \hspace{0.5pt} 95/46 \hspace{0.5pt} & 3991 & 4030 & 2145 & 2049 \\ \hline
          enqueue assumptions as set + \newline save literals after learnt units + \newline save literals from last invocation & 6138 & 115/29 & 6080 & \hspace{0.5pt} 92/42 \hspace{0.5pt} & 3989 & 4030 & 2149 & 2050 \\ \hline
        \end{tabular}
    \end{center}
    \caption{Number of \maxsat instances solved by MaxHS and RC2 using
      different extensions of the underlying \sat solver. The +/- column shows
      the number of instances gained/lost vs the original.}
    \label{fig:nsolved}
\end{figure}
\fi

>>>>

\title{Trail Saving on Backtrack}
\author{Randy Hickey \and Fahiem Bacchus}
\institute{Department of Computer Science, University of Toronto, Canada\\
  \email{rhickey@cs.toronto.edu, fbacchus@cs.toronto.edu}}

@@

In this paper we present a new trail saving method whereby we save
the backtracked part of the solver's trail and attempt to use that
information to make the solver's redescent more efficient. Unlike
\cbt, our trail saving method preserves the traditional invariants of
the SAT solver and its basic version is very simple to implement. It
allows the search to retain complete control over the order of
decisions, but helps make propagation faster. We develop some
enhancements to make the idea more effective, and demonstrate
experimentally that it performs as well as and often better than
chronological backtracking. We also show that with our enhancements we
are able to improve the performance of state-of-the-art solvers.

@@

This means that the decision levels on the trail are no longer
contiguous, as $\deepestLit$ has a different level than the other
literals at level $j$ (if $j\neq \btL$). This change has a number of
consequences for the \sat solver's operation, all of which were
described in \cite{DBLP:conf/lpar/JiangZ13}. M{\"{o}}hle and Biere
\cite{DBLP:conf/sat/MohleB19} showed that despite these consequences
partial backtracking can be made to preserve the soundness of a CDCL
solver.

@@

The second impact, changing variable scores, is partially addressed in
\cite{DBLP:conf/sat/MohleB19} who utilize the ideas of
\cite{DBLP:journals/jsat/TakRH11} to backtrack to a level where the
decisions would be unchanged. However, if the length of the backtrack
is greater than $100$ there could still be a divergence between the
variable decisions generated in standard backtracking and \cbt. An
argument is also given in \cite{DBLP:journals/jsat/TakRH11} that the
third impact, changing literal reasons, is not significant. However,
the experiments in \cite{DBLP:journals/jsat/TakRH11} were run before
good notions of clause quality were known
\cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results indicate
that once clause quality is accounted for, changing the literal
reasons can have a significant impact.

@@

@@

@@

In Figure~\ref{fig:cadicalNew} we used the newest version of cadical
(downloaded as of January 1, 2020) as the baseline solver, in
Figure~\ref{fig:cadicalOld} we used the version of cadical published
in \cite{DBLP:conf/sat/MohleB19} as the baseline, and in
Figure~\ref{fig:maple} we used MapleLCMDist \cite{DBLP:conf/ijcai/LuoLXML17,xiao2017maplelrb} as
the baseline. Each of the baselines were run with standard
non-chronological backtracking. We then refer to versions of each
solver with additional features implemented on top by adding
suffixes. "-chrono" refers to the solver with \cbt enabled (using the
solver's default settings), "-trail" refers to the baseline with plain
trail saving added (as described in Figure~\ref{fig:trailUse}),
"-trail-multipleBT" refers to the baseline with trail saving plus the
first enhancement of saving over multiple backtracks,
"-trail-multipleBT-lookahead" also adds the enhancement of lookahead
for conflicts by 2 decision levels, and
"-trail-multipleBT-lookahead-reason" also adds the final enhancement
to cease trail saving once a reason of "low quality" is reached. For
more details on the enhancements, please see section 4.2. The source
code for our implementation of trail saving can be obtained from the
web\footnote{https://github.com/rgh000/cadical-trail} or by downloading our submissions to the 2020 SAT Competition.

@@

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      cadical                  & 532          & 314 &  218  & 4025                             \\ \hline
      cadical-chrono          & 525          & 308 &  217  & 4086                             \\ \hline
      cadical-trail           & 529          & 310 &  \textbf{219}  & 4060                   \\ \hline
      cadical-trail-multipleBT & 530 & 313 & 217 & 3930 \\ \hline
      cadical-trail-multipleBT-lookahead & 531          & 313 &  218  & 4020                      \\ \hline
      cadical-trail-multipleBT-lookahead-reason & \textbf{538} & \textbf{319} & \textbf{219} & \textbf{3854} \\ \hline
    \end{tabular}
    \caption{Table of results for cadical, version pulled from github as of January 1, 2020.}
    \label{fig:cadicalNew}
\end{figure}

@@

@@

@@
